---
title: "Covid World Report"
author: "Bruno Zamorano"
date: "11/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
# Libraries
library(utils)
library(httr)
library(tidyverse)
library(scales)

# Style
theme_set(theme_minimal())
ggplot <- function(...) ggplot2::ggplot(...) + scale_color_brewer(palette="Set1")
blank_theme <- theme_minimal()+
  theme(
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=14, face="bold")
  )
my_percent <- label_percent( accuracy = 1)

# Download the data
setwd("~/Work/covid")

GET("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", 
    authenticate(":", ":", type="ntlm"),
    write_disk(tf <- "./covid_data.csv", overwrite = TRUE))

data <- read.csv2(tf, sep=",", header = TRUE)

# Replace some names to make plots nicer
data[data=="United_States_of_America"]<-"USA"
data[data=="United_Kingdom"]<-"UK"
data[data=="Democratic_Republic_of_the_Congo"]<-"DR_Congo"
data[data=="Bosnia_and_Herzegovina"]<-"Bosnia_Herz"

# # Population (from Wikipedia)
# pop <- read.csv("./population.csv")
# data<-merge(x=data, y=pop, by.x = "countriesAndTerritories",  by.y = "Country",all.x=TRUE)

# Fix the data first
data[[1]] <- as.POSIXct(strptime(data[[1]], "%d/%m/%Y"))
data <- data %>% arrange(desc(dateRep))
data[[10]] <- as.numeric(data[[10]])
colnames(data)[colnames(data) == "notification_rate_per_100000_population_14.days"] <- "incidence"

nrows <- nrow(data)

tot_sum <- integer(nrows)
death_sum <- integer(nrows)

countries <- unique(data$countriesAndTerritories)

for (country in countries) {
  irows <- (1:nrows)[data$countriesAndTerritories == country]
  dt <- data[irows,]

  n_rows <- nrow(dt)
  
  for (i in 1:n_rows ) {
    tot_sum[irows[i]] <- sum(dt$cases_weekly[i:n_rows])
    death_sum[irows[i]] <- sum(dt$deaths_weekly[i:n_rows])
  }
}

data['totCases'] <- tot_sum
data['totDeaths'] <- death_sum

spain <- data[data$countriesAndTerritories == "Spain",]

rm(dt, countries, country, death_sum, i, irows, n_rows, tf, tot_sum, nrows)

dd <- data%>%
  group_by(dateRep) %>%
  summarise(Cases = sum(totCases), Deaths = sum(totDeaths))

TotWorldCases <- max(dd$Cases)
TotWorldDeaths <- max(dd$Deaths)
LastDay <- max(dd$dateRep)
rm(dd)
```

## Overview up to `r toString(LastDay)`

This is the evolution of cases and fatalities in the World since the beginning of the pandemic. Note the logarithnmic y-axis:

```{r overview, echo=FALSE}
data%>%
  group_by(dateRep) %>%
  summarise(Cases = sum(totCases), Deaths = sum(totDeaths)) %>%
  ggplot(aes(x=dateRep)) +
  geom_line(aes(y=Cases), colour = "blue", size = 1.25) +
  geom_line(aes(y=Deaths), colour = "red", size = 1.25) +
  scale_y_log10(labels = comma_format(big.mark = " ")) +
  labs(x = "Date", y = "Total cases") +
  annotate("text", x=LastDay-86400*60, y=15000, label= paste("Total cases:", TotWorldCases), color = "blue")  +
  annotate("text", x=LastDay-86400*60, y=3000, label= paste("Total deaths:", TotWorldDeaths), color = "red")
```

The most affected country in total number of cases continues to be the United States.

```{r cases, echo=FALSE}
data %>%
  group_by(countriesAndTerritories) %>%
  summarise(Cases = max(totCases)) %>%
  arrange(desc(Cases)) %>%
  slice(1:15) %>%
  ggplot(aes(x = reorder(countriesAndTerritories, Cases), weight = Cases)) +
  geom_bar(aes(fill = Cases)) +
  scale_fill_gradient2( high = "darkblue") +
  theme_minimal() +
  guides(fill = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma_format(big.mark = " ")) +
  labs(x = "Country or territory", y = "Cases")
```

When scaled to the total population, small but highly developed territories tend to rank higher.

```{r cases scaled, echo=FALSE}
data %>%
  group_by(countriesAndTerritories) %>%
  summarise(Cases = max(totCases),
            Population = max(popData2019),
            Cases_rate = max(totCases) * 1.e3 / max(popData2019) ) %>%
  arrange(desc(Cases_rate)) %>%
  slice(1:15) %>%
  ggplot(aes(x = reorder(countriesAndTerritories, Cases_rate), weight = Cases_rate)) +
  geom_bar(fill = "orange", colour = "orange") +
  coord_flip() +
  labs(x = "Country or territory", y = "Cases per thousand inhabitants")
```

In terms of total deaths, the USA is also on top:

```{r deaths, echo=FALSE}
data %>%
  group_by(countriesAndTerritories) %>%
  summarise(Deaths = max(totDeaths)) %>%
  arrange(desc(Deaths)) %>%
  slice(1:15) %>%
  ggplot(aes(x = reorder(countriesAndTerritories, Deaths), weight = Deaths)) +
  geom_bar(aes(fill = Deaths)) +
  scale_fill_gradient2(high = "red") +
  theme_minimal() +
  guides(fill = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma_format(big.mark = " ")) +
  labs(x = "Country or territory", y = "Deaths")
```

But again, when scaled by population, smaller states appear higher on the list:

```{r death rate, echo=FALSE}
data %>%
  group_by(countriesAndTerritories) %>%
#  filter(max(popData2019) > 50000) %>%
#  filter(max(totDeaths) > 100) %>%
  summarise(Deaths = max(totDeaths),
            Population = max(popData2019),
            Death_rate = max(totDeaths) * 1.e6 / max(popData2019) ) %>%
  arrange(desc(Death_rate)) %>%
  slice(1:15) %>%
  ggplot(aes(x = reorder(countriesAndTerritories, Death_rate), weight = Death_rate)) +
  geom_bar(fill = "darkred", colour = "darkred") +
  coord_flip() +
  labs(x = "Country or territory", y = "Deaths per million inhabitants")
```

```{r prepare continents, include=FALSE}
data_continents <- data %>%
  group_by(continentExp, dateRep) %>%
  summarise(Deaths = sum(deaths_weekly),
            Cases = sum(cases_weekly))

nrows <- nrow(data_continents)
death_sum <- integer(nrows)
cases_sum <- integer(nrows)

continents <- unique(data_continents$continentExp)

for (continent in continents) {
  irows <- (1:nrows)[data_continents$continentExp == continent]
  dt <- data_continents[irows,]
  
  n_rows <- nrow(dt)
  
  for (i in 1:n_rows ) {
    death_sum[irows[i]] <- sum(dt$Deaths[1:i])
    cases_sum[irows[i]] <- sum(dt$Cases[1:i])
  }
}

data_continents['totDeath'] <- death_sum
data_continents['totCases'] <- cases_sum

rm(dt, cases_sum, continent, continents, death_sum, i, irows, n_rows, nrows)
```

### Continent analysis

The situation appears to remain under control in Oceania and, to a lesser extent, Asia. The trend in Africa is becoming worryingly steep again. The Americas and Europe appear to be entering a stabilising phase.

```{r continents timeline, echo=FALSE}
data_continents  %>%
  group_by(dateRep, continentExp) %>%
  filter(continentExp != "Other") %>%
  ggplot(aes(x = dateRep, y = totCases, colour = continentExp)) +
  geom_line(size=1) +
  scale_y_log10(labels = comma_format(big.mark = " ")) +
  labs(x = "Date", y = "Total cases", colour = "Continent")

data_continents  %>%
  group_by(dateRep, continentExp) %>%
  filter(continentExp != "Other") %>%
  ggplot(aes(x = dateRep, y = totDeath, colour = continentExp)) +
  geom_line(size=1) +
  scale_y_log10(labels = comma_format(big.mark = " ")) +
  labs(x = "Date", y = "Total deaths", colour = "Continent")
```
